---
title: "fastFMM: Functional mixed models with fast univariate inference"
output: rmarkdown::html_vignette
author: Gabriel Loewinger, Erjia Cui, Alison W Xin
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{fastFMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r preamble, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fastFMM)
```

## Introduction

`fastFMM` is a toolkit for fitting functional mixed models (FMMs) and implements the fast univariate inference (FUI) method proposed in [Cui et al. (2022)](https://doi.org/10.1080/10618600.2021.1950006). In this vignette, we provide a tutorial on how to use the `fui` function to fit non-concurrent and concurrent FMMs. 

### Installation

While `fastFMM` is available on CRAN, the development version of the package can be downloaded from [GitHub](https://github.com/gloewing/fastFMM):

```{r, eval = FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("gloewing/fastFMM", dependencies = T)
```

### Fast univariate inference (FUI)

Let $Y_{i, j} (s)$ be observations of multi-level/longitudinal functional data on the compact functional domain $\mathcal{S} = \{s_1, s_2, \dots, s_L\}$. Let $i = 1, 2, ..., I$ be the index of the subject and $j = 1, 2, ..., J_i$ be the index of longitudinal visit at time $t_{i, j}$. For each visit of each subject, we observe the fixed effects column vector $\mathbf{X}_{i, j} = [X_{i, j, 1}, X_{i, j, 2}, \dots, X_{i, j, p}]^T \in \mathbb{R}^p$ and the random effects column vector $\mathbf{Z}_{ij} = [Z_{i, j, 1}, Z_{i, j, 2}, ..., Z_{i, j, q}]^T \in \mathbb{R}^q$. Let $g(\cdot)$ be some pre-specified link function and $EF(\cdot)$ be some exponential family distribution. A longitudinal function-on-scalar regression model has the form

$$
\begin{aligned} 
  & Y_{ij}(s) \sim EF\{\mu_{ij}(s)\}, \\
  & g\{\mu_{ij}(s)\} = \eta_{ij}(s) = \boldsymbol{X}_{ij}^T\boldsymbol{\beta}(s) + \boldsymbol{Z}_{ij}^T\boldsymbol{u}_i(s),
\end{aligned} 
$$ 

referred to as a functional mixed model (FMM) in the functional data analysis (FDA) literature. Many statistical methods have been proposed to fit FMMs, and generally fall into two categories: joint methods and marginal methods. FUI is a marginal method that is computationally fast and achieves similar estimation accuracy compared with the state-of-the-art joint method, such as the `refund::pffr()` function.

FUI consists of the following three steps:

1. *Fit univariate models*. Fit separate linear mixed models at every point along the functional domain $\mathcal{S}$. That is, at each location $s_l \in \mathcal{S}, l = 1, 2, ..., L$,  we fit a pointwise generalized linear mixed model (GLMM) of the form
$$
\begin{aligned} 
  & Y_{i, j}(s_l) \sim EF\{\mu_{i,j}(s_l)\}, \\
  & g\{\mu_{i, j}(s_l)\} = \eta_{i,j}(s_l) = \boldsymbol{X}_{i,j}^T\boldsymbol{\beta}(s_l) + \boldsymbol{Z}_{i,j}^T\boldsymbol{u}_i(s_l),
\end{aligned}
$$ 
where $\boldsymbol{\beta}(s_l)$ is a $p \times 1$ dimensional vector of fixed effects and $\boldsymbol{u}_i(s_l)$ is a $q \times 1$ dimensional vector of random effects. Let $\boldsymbol{\tilde{\beta}}(s_1), \boldsymbol{\tilde{\beta}}(s_2), ..., \boldsymbol{\tilde{\beta}}(s_L)$ be the estimates of fixed effects from $L$ separate univariate GLMMs.

2. *Smooth coefficients*. Using penalized splines, aggregate and smooth the coefficient estimates $\boldsymbol{\tilde{\beta}}(s_1), \boldsymbol{\tilde{\beta}}(s_2), ..., \boldsymbol{\tilde{\beta}}(s_L)$ to produce estimates $\{\boldsymbol{\hat{\beta}}(s), s \in \mathcal{S}\}$. 
3. *Build pointwise and joint confidence intervals*. Combine the within-timepoint variance and the between-timepoint covariance to create confidence bands around the smoothed estimates from Step 2. For Gaussian data, we can obtain these estimates analytically. For other distributions, we implement cluster bootstrap. 

FUI decomposes the complex correlation structure into longitudinal and functional directions, allowing for a computationally efficient estimation procedure. For more details on the analytic and bootstrap approach, refer to [Cui et al. (2022)](https://doi.org/10.1080/10618600.2021.1950006).

### Additional references

We describe applications of `fastFMM` to photometry signal analysis in [Loewinger et al. (2024)](https://doi.org/10.7554/eLife.95802.2) and Xin et al. (2025). In Loewinger et al. (2024), we detailed advantages of functional mixed modeling over conventional photometry analysis methods, including the ability to generate hypothesis tests at every trial timepoint, incorporate the photometry signal for every animal and trial, capture nested effects, and compare temporal dynamics. In Xin et al. (2025), we focused on the extension of the FUI method to concurrent FMMs and described particular cases where concurrent FMMs produce more interpretable coefficient estimates.

## Reading data

The function `fastFMM::fui` requires that data be presented in a long format, i.e., with columns corresponding to locations on the functional domain $1, 2, \dots, L$ and with rows corresponding to longitudinal observations $t_{i, j}$. For example, within `fastFMM`, we provide the pre-formatted dataset `lick`, sourced from [Jeong et al. (2022)](https://doi.org/10.1126/science.abq6740). 

```{r read lick}
# Replace this with a package call
lick <- load(file = "../data/lick.rda", verbose = T)
```

### Data formatting

The columns `id, session, trial` correspond to trial-level covariates. The functional outcome of `photometry` is stored in the columns `photometry_1, photometry_2, ..., photometry_50` and the functional covariate `lick` is stored in the columns `lick_1, lick_2, ..., lick_50`. The `lick` dataset importable through `fastFMM` is a downsampled version of the original, containing half as many trials and half as many recording timepoints. 

We also provide the option to save functional outcomes and covariates as `AsIs` matrix columns. There are no particular differences between these two data structures. Below is an example of how an `AsIs` matrix column might be created.

```{r lick asis, eval = F}
# AX: fix this
library(dplyr)
lick_asis <- lick %>%
  mutate(
    photometry = matrix(dplyr::select(lick, photometry_1:photometry_43)))
  ) %>%
  select(-photometry_1:-photometry_50, -lick_1:-lick_43)
```

In the remainder of this tutorial, we will continue to use the original data frame `lick`. Below is an excerpt from `lick`, ignoring the columns containing scalar trial-specific covariates.

```{r lick colnames, echo = F}
knitr::kable(head(lick[, 1:12] %>% select(-lick_rate_050:-iri)))
```

The data frame `lick` has $N = \sum_{i = 1}^I n_i$ rows, where $n_i$ is the number of repeated measures observations of subject/cluster $i$. The first three columns of the dataset include the covariates subject ID (`id`), session number (`session`), and trial number (`trial`). Each of the remaining columns are observations of our functional outcome named $[\texttt{photometry_1}, \; \texttt{photometry_2}, \dots, \; \texttt{photometry_L}]$, where $L$ is the size of the functional domain. The prefix `photometry_` is arbitrary and can be replaced with any convenient outcome name. However, because `fastFMM::fui` uses initial string matching to find functional outcome and covariate columns, the identifying prefix substring should be unique and not shared with any other columns. Furthermore, `fastFMM::fui` assumes any functional columns are ordered from left to right. 

## Non-concurrent FMM fitting

In this dataset, we will use `lick_rate_NNN` as an example trial-specific behavioral summary. The covariate `lick_rate_NNN` is calculated as the average licking rate in the `NNN` hundredths of seconds after the reward is delivered. E.g., `lick_rate_050` is the average licking rate 0.5 seconds after reward delivery. We provide calculated lick rates for four different time periods (0.5, 1.0, 1.5, and 2.0 seconds). 

```{r lick cov names, echo = F}
knitr::kable(lick[1:5, 1:7])
```

Creating a FLMM follows the general formulat

```{r}
# AX: fix printing of test message
lick_prob_050 <- fastFMM::fui(
  photometry ~ lick_rate_050 + trial + session + (lick_rate_050|id), 
  data = lick
)
```


## Concurrent FMM fitting
